Open-Set Domain Adaptation through Self-Supervision

### 2022-01-21 Report/Paper_OpenSet.tex
Report - Completed ablation study

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}

@@

Stage I of the simplified ROS method involves training both an object classifier and a rotation recognition task on data from the source domain, as shown on the left side of Figure \ref{fig:ROS}, where $D_s$ is the source domain dataset, $E$ is the encoder, $C_1$ is the object classifier and $R_1$ is the rotation classifier. The object prediction is based on the features of the original source samples, whereas the rotation prediction is based on the concatenated features of the original and rotated samples. The object classification and rotation recognition tasks are trained simultaneously to minimize the total loss objective function given by:
\begin{equation}
  L_{tot} = L_{cls} + \alpha_1 L_{rot} ,
  \label{eq:totalloss}
\end{equation}
where $L_{cls}$ is the loss from the object classifier, $L_{rot}$ is the loss from the rotation classifier, and $\alpha_1$ is the weight assigned to the rotation recognition task. The value of $\alpha_1$ is tuned according to the performance of the network, as detailed in Section \ref{sec:ablation}.

@@

The precision of the normality score is then evaluated using the AUROC (Area Under Receiver Operating Characteristics curve) metric. The AUROC metric reduces analysis of the ROC curve to a single scalar value which can be used to compare the performance of classifiers \cite{Fawcett2006}. The higher the AUROC score the better the classifier overall, with scores $> 0.5$ showing an improvement over random guessing.

If the AUROC is $> 0.5$ then the normality score can be used to conduct the known/unknown separation subject to a selected threshold, i.e. if the normality score is above the threshold the sample is considered as known and added to $D^{knw}_t$, otherwise it is considered as unknown and added to $D_t^{unk}$. The threshold value is also tuned according to the performance of the network, as detailed in Section \ref{sec:ablation}.

\subsection{Stage II - domain alignment}

In stage II of the simplified ROS method the unknown dataset created from the target domain $D_t^{unk}$ is combined with the samples from the source domain $D_s$, with the unknown samples representing an additional `unknown' class. The object classifier $C_2$ is thus trained to recognise samples belonging to the unknown class. Concurrently, the known target samples  $D^{knw}_t$ are used for the source-target adaptation by training the rotation recognition task $R_2$. The method is shown on the right side of Figure \ref{fig:ROS}. As in stage I, the object classification and rotation recognition tasks are trained simultaneously to minimize the total loss objective function given by:
\begin{equation}
  L_{tot} = L_{cls} + \alpha_2 L_{rot} ,
  \label{eq:totalloss}
\end{equation}
where $\alpha_2$ is the weight assigned to the rotation recognition task of stage II. The value of $\alpha_2$ is also selected to provide the best performance, detailed further in Section \ref{sec:ablation}.

@@

- the network used

@@

- ablation study: hyperparameter tuning of weights and threshold value, include graphs or tables of values

>>>>

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage[export]{adjustbox}
\usepackage{multirow}

@@

Stage 1 of the simplified ROS method involves training both an object classifier and a rotation recognition task on data from the source domain, as shown on the left side of Figure \ref{fig:ROS}, where $D_s$ is the source domain dataset, $E$ is the encoder, $C_1$ is the object classifier and $R_1$ is the rotation classifier. The object prediction is based on the features of the original source samples, whereas the rotation prediction is based on the concatenated features of the original and rotated samples. The object classification and rotation recognition tasks are trained simultaneously to minimize the total loss objective function given by:
\begin{equation}
  L_{tot} = L_{cls} + \alpha_1 L_{rot} ,
  \label{eq:totalloss}
\end{equation}
where $L_{cls}$ is the loss from the object classifier, $L_{rot}$ is the loss from the rotation classifier, and $\alpha_1$ is the weight assigned to the rotation recognition task. The value of $\alpha_1$ is tuned according to the performance of the network, as detailed in Section \ref{sec:ablation}.

@@

The precision of the normality score is then evaluated using the AUC (area under receiver operating characteristics (ROC) curve) metric. The AUC metric reduces analysis of the ROC curve to a single scalar value which can be used to compare the performance of classifiers \cite{Fawcett2006}. The higher the AUC score the better the classifier overall, with scores $> 0.5$ showing an improvement over random guessing.

If the AUC is $> 0.5$ then the normality score can be used to conduct the known/unknown separation subject to a selected threshold, i.e. if the normality score is above the threshold the sample is considered as known and added to $D^{knw}_t$, otherwise it is considered as unknown and added to $D_t^{unk}$. The threshold value is also tuned according to the performance of the network, as detailed in Section \ref{sec:ablation}.

\subsection{Stage II - domain alignment}
\label{sec:stage2}

In stage 2 of the simplified ROS method the unknown dataset created from the target domain $D_t^{unk}$ is combined with the samples from the source domain $D_s$, with the unknown samples representing an additional `unknown' class. The object classifier $C_2$ is thus trained to recognise samples belonging to the unknown class. Concurrently, the known target samples  $D^{knw}_t$ are used for the source-target adaptation by training the rotation recognition task $R_2$. The method is shown on the right side of Figure \ref{fig:ROS}. As in stage I, the object classification and rotation recognition tasks are trained simultaneously to minimize the total loss objective function given by:
\begin{equation}
  L_{tot} = L_{cls} + \alpha_2 L_{rot} ,
  \label{eq:totalloss}
\end{equation}
where $\alpha_2$ is the weight assigned to the rotation recognition task of stage 2. The value of $\alpha_2$ is also selected to provide the best performance, detailed further in Section \ref{sec:ablation}.

@@

The proposed method was tested on the \textit{Office-Home} dataset \cite{OfficeHome}. This dataset consists of four separate domains for image classification: Art, Clipart, Product and Real World. An example of the difference between images in these domains is shown in Figure \ref{fig:domains}. Each domain included two sets of images, one when behaving as the source domain and another as the target, where the target domain included images from a number of categories that were non existant in the source domain set.

\begin{figure}[!htbp]
  \centering
  \begin{adjustbox}{minipage=\linewidth,scale=0.8}
  \begin{subfigure}[b]{0.45\columnwidth}
    \centering
    \includegraphics[width=\linewidth]{Figures/Art_Radio.jpg}
    \caption{Art}
    \label{fig:short-a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\columnwidth}
\centering
    \includegraphics[width=\linewidth]{Figures/Clipart_Radio.jpg}
    \caption{Clipart}
    \label{fig:short-b}
  \end{subfigure}
\vskip\baselineskip
  \begin{subfigure}[b]{0.45\columnwidth}
\centering
    \includegraphics[width=\linewidth]{Figures/Product_Radio.jpg}
    \caption{Product}
    \label{fig:short-c}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.45\columnwidth}
\centering
    \includegraphics[width=\linewidth]{Figures/RealWorld_Radio.jpg}
    \caption{Real World}
    \label{fig:short-d}
  \end{subfigure}
\end{adjustbox}
  \caption{Domains in the Office-Home dataset \cite{OfficeHome}.}
  \label{fig:domains}
\end{figure}

The experiments were conducted using a ResNet18 convolutional neural network architecture, employing a learning rate of 0.001, a batch size of 128. Image data augmentation techniques were also emploted to improve performance of the model, these included random resized crop, color jitter and random grayscale.

@@

The simplified ROS method includes three hyperparameters that need to be tuned according to the network: the weight assigned to the rotation recognition tasks in stage 1 ($\alpha_1$); the threshold of normality score for considering a target sample as belong to a known category; and the weight assigned to the rotation recognition tasks in stage 2 ($\alpha_2$). The $\alpha_1$ parameter of stage 1 was tuned first. Figure \ref{fig:ClassAcc} shows the class accuracy of stage 1 when trained on the \textit{Art} source domain for a number of epochs and different $\alpha_1$ values, and Figure \ref{fig:RotAcc} shows the corresponding rotation recognition accuracy. As expected the accuracy of both increases with increasing number of epochs, however lower weights take much longer to converge for the rotation recognition tasks. Since the two classifiers are trained simultaneously a balance needs to be sought to reduce over-fitting of either task.

\begin{figure}[!htb]
  \centering
   \includegraphics[width=0.95\linewidth]{Figures/ClassAcc.png}
   \caption{Class accuracy during stage 1 training on the Art domain for various weights.}
   \label{fig:ClassAcc}
\end{figure}

\begin{figure}[!htb]
  \centering
   \includegraphics[width=0.95\linewidth]{Figures/RotAcc.png}
   \caption{Rotation recognition accuracy during stage 1 training on the Art domain for various weights.}
   \label{fig:RotAcc}
\end{figure}

The AUC metric was then used to compare the performance of the various weights. The AUC value for the network with \textit{Art} as the source domain and \textit{Clipart} as the target domain is shown in Table \ref{tab:auc}. The highest AUC measured was 0.5578 for a weight of 10 at 40 epochs, therefore 10 was chosen at the values for $\alpha_1$ in stage 1.

\begin{table}[!htb]
  \centering
  \begin{tabular}{@{}|l|ccccc|@{}}
    \hline
     & \multicolumn{5}{c|}{\textbf{Weight ( $\alpha_1$)}}\\
 \textbf{Epoch} & 0.1 & 0.5 & 1 & 3 & 10 \\
    \hline 
    40 & 0.5293 & 0.4944 & 0.5245 & 0.5492 & 0.5578 \\
   80 & 0.5203 & 0.4873 & 0.5312 & 0.5397 & 0.5548 \\
    \hline
  \end{tabular}
  \caption{AUC for Art -- Clipart}
  \label{tab:auc}
\end{table}

The next step involed tuning the threshold. The threshold should be chosen such that the number of samples recognised as known is similar to the number of samples that are expected to be recognised as known. Considering again the \textit{Art} as source and \textit{Clipart} as target domain combination as an example, the total number of samples in the \textit{Clipart} target domain is 4365, and 3064 of those samples share the same category as those in the \textit{Art} source domain. The model was thus evaluated with different thresholds and the number of samples identified as known analysed. Table \ref{tab:threshold} shows the results of the analysis on the \textit{Art}--\textit{Clipart} domain combination. A threshold of 0.75 results in the number of known samples being most similar to the expected number of known samples of 3064, and so 0.75 was chosen as the threshold for that combination.

\begin{table}[!htb]
  \centering
  \begin{tabular}{@{}|cc|@{}}
    \hline
 \textbf{Threshold} & \textbf{Known Samples} \\
    \hline 
 0.3 & 3778 \\
 0.5 & 3486 \\
 0.7 & 3123 \\
 0.75 & 3019 \\
 0.8 & 2872 \\
    \hline
  \end{tabular}
  \caption{Samples recognised as known for various thresholds for the Art -- Clipart domain combination}
  \label{tab:threshold}
\end{table}

The final step of the ablation study was to tune the stage 2 weight ($\alpha_2$). The evaluation metrics for stage 2 involve the harmonic mean between the accuracy of the known classes and the abiility of the classifier to recognize the unknown category, as discussed in \ref{sec:stage2}. The results of stage 2 for the \textit{Art}--\textit{Clipart} domain combination are shown in Table \ref{tab:tuneHOS}. The highest harmonic mean value is 0.4399, for a weight of 3 at 20 epochs.

\begin{table}[!htb]
  \centering
  \begin{tabular}{@{}|ccccc|@{}}
    \hline
 \textbf{Weight ($\alpha_2$}) & \textbf{Epoch} & \textbf{OS$^*$} & \textbf{UNK} & \textbf{HOS} \\
    \hline 
 0.01 & 20 & 0.3026 & 0.5778 & 0.3972 \\
 & 40 & 0.3008 & 0.5743 & 0.3948 \\ \hline
0.1 & 20 & 0.3091 & 0.5896 & 0.4056 \\
 & 40 & 0.3257 & 0.5601 & 0.4119 \\\hline
0.5 & 20 & 0.3280 & 0.5731 & 0.4172 \\
 & 40 & 0.3404 & 0.5495 & 0.4204 \\\hline
3 & 20 & 0.3542 & 0.5802 & 0.4399 \\
 & 40 & 0.3690 & 0.5000 & 0.4246 \\\hline
10 & 20 & 0.3114 & 0.6156 & 0.4136 \\
 & 40 & 0.3367 & 0.5743 & 0.4246 \\
    \hline
  \end{tabular}
  \caption{Stage 2 evaluation metrics for the Art -- Clipart domain combination}
  \label{tab:tuneHOS}
\end{table}

Similar analyses were also conducted for the 11 other possible domain combinations, the results of which can be found in the supplementary material.
